---
title: "Predicting Leave of Employees"
author: "Egor Gazarov, William Handjaja, Dmitry Kalachev, Nathan Kung (INSEAD MBA Class of July 2017); Jorge Bravo"
date: "January 31, 2017"
output: 
 html_document:
    css: Styles/default.css
    theme: paper
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r , echo=TRUE, message=FALSE, warning=FALSE}
# basic preparatory commands

# clearing the working environment
rm(list = ls())

# loading or installing packages used in the project
# make sure to list all required packages in ./R/library.R
source("./R/library.R")

# loading the dataset for the project
HR <- read.csv("./Data/HR_dataset.csv", sep=",", dec=".")

```

# Introduction

## Brief description of the project

This project has been done by a group of students of INSEAD course 'Big Data & Analytics for Business' (Professor [Theos Evgeniou](https://www.insead.edu/faculty-research/faculty/theodoros-evgeniou)) in January 2017. Using techniques learned in the course we will analyze dataset describing employees of an organization and build a model to predict if employee will leave a company in near future.


## Project process and document structure

Both project work and this report are structured according to following process (based on Cross Industry Standard Process for Data Mining):

**1. Business understanding.** Identification of business question to be answered by model prepared during the project.

**2. Data undestanding.** Initial analysis of the dataset used in the project using basic descriptive statistics and visualizations.

**3. Data preparation.** Preparation of the dataset for subsequent modeling.

+ Check for missing and empty values, exclusion of corresponding observations.
+ Check for outliers, decision on their participation in analysis.
+ Conversion of non-numerical attributes to numerical dummy variables.
+ Range normalization of all attributes (z-score normalization).
+ Dimensionality reduction (necessity to be discussed).
+ Separation of dataset into training and test.

**4. Modelling.** Building the predictive model based on training dataset with target attribute answering the chosen business question. We use classification trees methodology.

+ Decision on predictive quality of model to be considered acceptable.
+ Selection of classification tree algorithm and R package to be used.
+ Selection of explanatory variables.
+ Application of algorithm to training dataset and getting the model.
+ Generation of text and graphical visualizations of decision trees, checking for business sense.
+ Preliminary evaluation of the model using confusion matrix and test dataset. Decision to proceed (if evaluation results are satisfactory) or repeating the modelling process.

**5. Evaluation.** Check of models predictive accuracy on test dataset.

+ Generation of confusion matrix using built model and test dataset.
+ Plotting receiver operating characteristic curve for the model and test dataset.
+ Comparing evaluation results with target characteristics.
+ Decision on accepting the model or repeating the modelling process.

**6. Deployment.** Conclusion on project results and description of potential deployment of built model in practice. Remarks about potential ways to improve process and model of this project.


# Business Understanding

People are the most important resource for a lot of companies. And employees turnover has always been one of major HR issues. Retaining workers is becoming a balancing act between hard data research and a human touch. It’s more important than ever to complement intuition with statistical analysis (more on this available, for example on [INSEAD Knowledge](http://knowledge.insead.edu/blog/insead-blog/the-art-of-keeping-employees-from-leaving-3896).

Business problem that this project intends to solve is **prediction of employees to leave in near future using attributes known about each of employees**.

# Data Understanding

## Source of data

For this project we use dataset "" published by Ludovic Benistant on kaggle ([source](https://www.kaggle.com/ludobenistant/hr-analytics/)) under CC BY-SA 4.0 License. 

## Dataset size and variables

Dataset contains `r nrow(HR)` observations. Each observation represents information about one employee across `r ncol(HR)` variables. Variables are (we indicate code of variable in dataset in parentheses):

+ Employee satisfaction level *(satisfaction_level)*
+ Last evaluation *(last_evaluation)*
+ Number of projects *(number_project)*
+ Average monthly hours *(average_montly_hours)*
+ Time spent at the company *(time_spend_company)*
+ Whether they have had a work accident *(Work_accident)*
+ Whether they have had a promotion in the last 5 years *(promotion_last_5years)*
+ Department *(sales)*
+ Salary *(salary)*
+ Whether the employee has left, the target variable in this project *(left)*

Below is summary of project dataset structure.

```{r}
str(HR)
```

## Distributions of variables {.tabset}

First, let's generate summary report for all variables.

```{r , echo=TRUE, message=FALSE, warning=FALSE}
summary(HR)
```

We will visualize distributions of variables separately for people who left the company and who stayed to be able to see differences between these groups of people on charts before doing the actual modelling.

```{r , echo=TRUE, message=FALSE, warning=FALSE}
HR.left <- subset(HR, left == 1)
HR.stayed <- subset(HR, left == 0)
```

There are `r nrow(HR.left)` observations of employees who left the company and `r nrow(HR.stayed)` observations of employees who stayed.

### Satisfaction level

```{r , echo=TRUE, message=FALSE, warning=FALSE}
satisfaction.left <- ggplot(HR.left, aes(HR.left$satisfaction_level)) + geom_density(kernel = "gaussian", fill = '#B8274C', alpha = 0.3) + labs(x = "Satisfaction level of employees who left") + xlim(0, 1)
satisfaction.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
satisfaction.stayed <-ggplot(HR.stayed, aes(HR.stayed$satisfaction_level)) + geom_density(kernel = "gaussian", fill = '#006E51', alpha = 0.3) + labs(x = "Satisfaction level of employees who stayed") + xlim(0, 1)
satisfaction.stayed
```

### Last evaluation

```{r , echo=TRUE, message=FALSE, warning=FALSE}
evaluation.left <- ggplot(HR.left, aes(HR.left$last_evaluation)) + geom_density(kernel = "gaussian", fill = '#B8274C', alpha = 0.3) + labs(x = "Last evaluation of employees who left") + xlim(0, 1)
evaluation.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
evaluation.stayed <-ggplot(HR.stayed, aes(HR.stayed$last_evaluation)) + geom_density(kernel = "gaussian", fill = '#006E51', alpha = 0.3) + labs(x = "Last evaluation of employees who stayed") + xlim(0, 1)
evaluation.stayed
```

### Number of projects

```{r , echo=TRUE, message=FALSE, warning=FALSE}
projects.left <- ggplot(HR.left, aes(HR.left$number_project)) + geom_histogram(fill = '#B8274C', binwidth = .5) + scale_x_continuous(breaks = seq(0,max(max(HR.left$number_project),max(HR.stayed$number_project)),by = 1), limits = c(0, max(max(HR.left$number_project),max(HR.stayed$number_project))+1)) + labs(x = "Number of projects of employees who left")
projects.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
projects.stayed <- ggplot(HR.stayed, aes(HR.stayed$number_project)) + geom_histogram(fill = '#006E51', binwidth = .5) + scale_x_continuous(breaks = seq(0,max(max(HR.left$number_project),max(HR.stayed$number_project)),by = 1), limits = c(0, max(max(HR.left$number_project),max(HR.stayed$number_project))+1)) + labs(x = "Number of projects of employees who stayed")
projects.stayed
```

### Monthly hours

```{r , echo=TRUE, message=FALSE, warning=FALSE}
hours.left <- ggplot(HR.left, aes(HR.left$average_montly_hours)) + geom_density(kernel = "gaussian", fill = '#B8274C', alpha = 0.3) + labs(x = "Average monthly hours of employees who left") + xlim(min( min(HR.left$average_montly_hours), min(HR.stayed$average_montly_hours)), max( max(HR.left$average_montly_hours), max(HR.stayed$average_montly_hours)))
hours.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
hours.stayed <- ggplot(HR.stayed, aes(HR.stayed$average_montly_hours)) + geom_density(kernel = "gaussian", fill = '#006E51', alpha = 0.3) + labs(x = "Average monthly hours of employees who stayed") + xlim(min( min(HR.left$average_montly_hours), min(HR.stayed$average_montly_hours)), max( max(HR.left$average_montly_hours), max(HR.stayed$average_montly_hours)))
hours.stayed
```

### Time at company

```{r , echo=TRUE, message=FALSE, warning=FALSE}
time.left <- ggplot(HR.left, aes(HR.left$time_spend_company)) + geom_histogram(fill = '#B8274C', binwidth = .5) + scale_x_continuous(breaks = seq(0,max(max(HR.left$time_spend_company),max(HR.stayed$time_spend_company)),by = 1), limits = c(0, max(max(HR.left$time_spend_company),max(HR.stayed$time_spend_company))+1)) + labs(x = "Time spent at the company (in years) of employees who left")
time.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
time.stayed <- ggplot(HR.stayed, aes(HR.stayed$time_spend_company)) + geom_histogram(fill = '#006E51', binwidth = .5) + scale_x_continuous(breaks = seq(0,max(max(HR.left$time_spend_company),max(HR.stayed$time_spend_company)),by = 1), limits = c(0, max(max(HR.left$time_spend_company),max(HR.stayed$time_spend_company))+1)) + labs(x = "Time spent at the company (in years) of employees who stayed")
time.stayed
```

### Work accidents 

```{r , echo=TRUE, message=FALSE, warning=FALSE}
accidents.left <- ggplot(HR.left, aes(HR.left$Work_accident)) + geom_histogram(fill = '#B8274C', binwidth = .5) + labs(x = "Number of employees who experienced and did not experience work accident of employees who left")
accidents.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
accidents.stayed <- ggplot(HR.stayed, aes(HR.stayed$Work_accident)) + geom_histogram(fill = '#006E51', binwidth = .5) + labs(x = "Number of employees who experienced and did not experience work accident of employees who stayed")
accidents.stayed
```

### Promotions 

```{r , echo=TRUE, message=FALSE, warning=FALSE}
promotion.left <- ggplot(HR.left, aes(HR.left$promotion_last_5years)) + geom_histogram(fill = '#B8274C', binwidth = .5) + labs(x = "Number of employees who were or were not promoted during last 5 years of employees who left")
promotion.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
promotion.stayed <- ggplot(HR.stayed, aes(HR.stayed$promotion_last_5years)) + geom_histogram(fill = '#006E51', binwidth = .5) + labs(x = "Number of employees who were or were not promoted during last 5 years of employees who stayed")
promotion.stayed
```

### Department

```{r , echo=TRUE, message=FALSE, warning=FALSE}
Departments <- c("sales", "technical", "support", "IT", "product_mng", "marketing", "RandD", "accounting", "hr","management")
department.left <- ggplot(HR.left, aes(HR.left$sales)) + geom_bar(stat = "count", fill = '#B8274C') + labs(x = "Department of employees who left") + scale_x_discrete(limits = Departments)
department.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
department.stayed <- ggplot(HR.stayed, aes(HR.stayed$sales)) + geom_bar(stat = "count", fill = '#006E51') + labs(x = "Department of employees who stayed") + scale_x_discrete(limits = Departments)
department.stayed
```

### Salary

```{r , echo=TRUE, message=FALSE, warning=FALSE}
salary.left <- ggplot(HR.left, aes(HR.left$salary)) + geom_bar(stat = "count", fill = '#B8274C') + labs(x = "Salary level of employees who left") + scale_x_discrete(limits = c("low", "medium", "high"))
salary.left
```

```{r , echo=TRUE, message=FALSE, warning=FALSE}
salary.stayed <- ggplot(HR.stayed, aes(HR.stayed$salary)) + geom_bar(stat = "count", fill = '#006E51') + labs(x = "Salary level of employees who stayed") + scale_x_discrete(limits = c("low", "medium", "high"))
salary.stayed
```

## Correlations

Let's check correlations between variables.

```{r}
CorTable <- cor(HR[,1:8])
corrplot(CorTable, method = "number", type= "upper")
```

`r if (max(CorTable [CorTable != 1]) < 0.7) { "Analysis shows that there are no strong correlation (0.7 or avobe) between any pair of numeric variables."} else "Analysis shows that strong correlation (0.7 or above) between numeric variables of the set is observed."`


# Data preparation

Work in progress


# Modelling

Work in progress


# Evaluation

Work in progress


# Deployment

Work in progress